---
title: "WaveformInCSV"
author: "Rob Donald"
date: "`r format(Sys.time(), '%A %d %B %Y, %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Introduction

This analysis will look at extracting waveform data that is encoded in the 
columns of a .csv file. 

Each row of the .csv file will contain the complete 11 point waveform along
with meta data that describes the conditions at the time the waveform was captured.

This is an example of a complex data structure contained within a seemingly
simple .csv file. This becomes a lot trickier than it looks.

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE,fig.height=8,fig.width=12)
knitr::opts_chunk$set(echo = TRUE)
#options(width=1500)
```

## Libraries

```{r library_setup}
suppressMessages({suppressWarnings({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(ggplot2)
  library(gridExtra)
  library(data.table)
  
  library(RobsRUtils)
  library(futile.logger)
})})
```

# Generate Data

```{r}

all.traces <- NULL

all.decay.profiles <- c(5,10,15,20)

for(decay.profile in all.decay.profiles)
{
    time.ms <- seq(0,100,by=10)
    response <- exp(-(time.ms/decay.profile))
    
    decay.setting <- rep(decay.profile,length(time.ms))
    waveform.df <- data_frame(time.ms,response,decay.setting)
    
    if(is.null(all.traces))
    {
       all.traces <- waveform.df
    }
    else
    {
       all.traces <- bind_rows(all.traces,waveform.df) 
    }
}

```

We now add in some batch and experiment day information

```{r}
all.traces$batch <- ifelse(all.traces$decay.setting > 10,1234,5678)
all.traces$exp.day <- ifelse(all.traces$decay.setting %% 10 == 0,'Day 1','Day 2')
```


```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
print(p)
```



```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response [Panel: batch]'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
p <- p + facet_grid(. ~ batch)
print(p)
```


```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response [Panel: batch, experiment day]'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
p <- p + facet_grid(exp.day ~ batch)
print(p)
```

Now let's save that out in a .csv format where each row is the trace from an 
experimental row along with the meta data (in this case the decay setting, 
batch and experiment day) from that run.

We have four experimental runs:

+ Day 1
    + Batch 1234
        + Decay setting 20
        
+ Day 1
    + Batch 5678
        + Decay setting 10
        
+ Day 2
    + Batch 1234
        + Decay setting 15
        
+ Day 2
    + Batch 5678
        + Decay setting 5        

So this means we will have four rows in our .csv file.

Let's pull out each run's results into an R object. 

```{r}
d1.b1234.ds20 <- filter(all.traces, exp.day == 'Day 1', batch == 1234, decay.setting == 20)
d1.b5678.ds10 <- filter(all.traces, exp.day == 'Day 1', batch == 5678, decay.setting == 10)

d2.b1234.ds15 <- filter(all.traces, exp.day == 'Day 2', batch == 1234, decay.setting == 15)
d2.b5678.ds05 <- filter(all.traces, exp.day == 'Day 2', batch == 5678, decay.setting == 5)
```

For each of these objects we have 11 rows of 5 variables. We need to flatten this data
into a single row. In actual fact we have 3 bits of meta data (decay setting, 
batch and experiment day) and the 11 reading from the waveform 0 to 100 ms. We are going
to construct 14 column row which we can write out in .csv format.

First we collect the waveform data into a single numeric vector.

```{r}
exp.df <- d1.b1234.ds20
wf <- NULL
num.time.pts <- nrow(exp.df)

for (count in 1:num.time.pts)
{
    wf <- c(wf,exp.df$response[count])
}
```

Now we make another vector with the meta data. We only need to grab the first 
element of the required vector.

```{r}
meta.vec <- c(exp.df$exp.day[1],exp.df$batch[1],exp.df$decay.setting[1])
```

Then we can stick these two vectors together and give them names.

```{r}
full.row <- c(meta.vec,wf)
names(full.row) <- c('ExpDay','Batch','DecaySetting'
                        ,'t=0','t=10','t=20','t=30','t=40','t=50'
                        ,'t=60','t=70','t=80','t=90','t=100')

full.row
```
Let's put the above techniques into a function

```{r}
build_WF_Row <- function(exp.df)
{
    wf <- NULL
    num.time.pts <- nrow(exp.df)
    
    for (count in 1:num.time.pts)
    {
        wf <- c(wf,exp.df$response[count])
    }
    
    meta.vec <- c(exp.df$exp.day[1],exp.df$batch[1],exp.df$decay.setting[1])

    full.row <- c(meta.vec,wf)
    
    # Having got a full row vector we want it as a data frame with
    # a column for each element. The as.is stops the 'Day x' becoming a factor.
    full.row.df <- data.frame(lapply(full.row, type.convert,as.is=TRUE), stringsAsFactors=FALSE)
    
    # We then set the names including the slightly odd t= pattern.
    names(full.row.df) <- c('ExpDay','Batch','DecaySetting'
                        ,'t=0','t=10','t=20','t=30','t=40','t=50'
                        ,'t=60','t=70','t=80','t=90','t=100')
    
    return(full.row.df)
}
```


Now we'll build up a 4 row object with the data from the experiment. We would do this
in a for() loop or using lapply but for this example we'll do this manually.

```{r}
full.exp.list <- list()
full.exp.list[[1]] <- build_WF_Row(d1.b1234.ds20)
full.exp.list[[2]] <- build_WF_Row(d1.b5678.ds10)
full.exp.list[[3]] <- build_WF_Row(d2.b1234.ds15)
full.exp.list[[4]] <- build_WF_Row(d2.b5678.ds05)
```
We have a list of data_frames but what we want it a single data_frame. 
Use rbindlist from th data.table package to achieve this. For some 

```{r}
full.exp.df <- rbindlist(full.exp.list)
```


Let's check that looks as we expect.


```{r}
full.exp.df[,c('ExpDay','Batch','DecaySetting','t=0','t=10','t=90','t=100')]
```

# Export the data
We can now write this out to a .csv file so that you can prove to yourself
that it is what you would expect.

I'll assume that you current directory is the project dir.

```{r}
exp.file.name <- 'ExperimentFullData.csv'
write.csv(full.exp.df,file = exp.file.name, row.names = FALSE )
```

# Import the data
What we want to do now is show how you can read in a file in this format
and construct an object that ggplot can use to produce graphs like those above.

```{r}
raw.exp.data <- read_csv(file = exp.file.name)
```

# Construct data for ggplot()

What we want is a function that can take a row of meta data *and* waveform data
and return a data_frame that can we can use in a bind_rows construct to form an
object for ggplot().

```{r}
buildWaveformAndMetaData<-function(single.row)
{
    # Our first task is to split out the waveform from the metadata.
    
    wf <- select(single.row,contains('t='))
    col.names <- names(wf)
    
    time.ms <- rep(NaN,length(col.names))
    response <- rep(NaN,length(col.names))
    waveform.df <- data_frame(time.ms,response)
    for(count in 1:length(col.names))
    {
      waveform.df$time.ms[count] <- gsub('t=','',col.names[count])
      waveform.df$response[count] <- as.numeric(wf[count])
    }
    
    waveform.df$time.ms <- as.numeric(waveform.df$time.ms)

    meta.data <- select(single.row,-contains('t='))
    col.names <- names(meta.data)
    
    for(column.name in col.names)
    {
        waveform.df[[column.name]] <- rep(meta.data[[column.name]],nrow(waveform.df))
    }
    
    return(waveform.df)
}
```

Let's test this function

```{r}
single.row <- raw.exp.data[1,]

one.row <- buildWaveformAndMetaData(single.row)
one.row
```

Now do it for the whole .csv file.

```{r}

all.experimental.data <- NULL
for(count in 1:nrow(raw.exp.data))
{
    single.row <- raw.exp.data[count,]

    one.row.df <- buildWaveformAndMetaData(single.row)
    
    if(is.null(all.experimental.data))
    {
        all.experimental.data <- one.row.df 
    }
    else
    {
        all.experimental.data <- bind_rows(all.experimental.data,one.row.df)
    }
    
}


```


```{r}
p <- ggplot(data=all.experimental.data,aes(x=time.ms,y=response,colour = as.factor(DecaySetting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experimental Response'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
print(p)
```


```{r}
p <- ggplot(data=all.experimental.data,aes(x=time.ms,y=response,colour = as.factor(DecaySetting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experimental Response [Panel: batch, experiment day]'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
p <- p + facet_grid(ExpDay ~ Batch)
print(p)
```
