---
title: "WaveformInCSV"
author: "Rob Donald"
date: "`r format(Sys.time(), '%A %d %B %Y, %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Introduction

This analysis will look at extracting waveform data that is encoded in the 
columns of a .csv file. 

Each row of the .csv file will contain the complete 11 point waveform along
with meta data that describes the conditions at the time the waveform was captured.

This is an example of a complex data structure contained within a seemingly
simple .csv file. This becomes a lot trickier than it looks.

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE,fig.height=8,fig.width=12)
knitr::opts_chunk$set(echo = TRUE)
#options(width=1500)
```

## Libraries

```{r library_setup}
suppressMessages({suppressWarnings({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(gridExtra)
  library(data.table)
  
  library(RobsRUtils)
  library(futile.logger)
})})
```

# Generate Data

```{r}

all.traces <- NULL

all.decay.profiles <- c(5,10,15,20)

for(decay.profile in all.decay.profiles)
{
    time.ms <- seq(0,100,by=10)
    response <- exp(-(time.ms/decay.profile))
    
    decay.setting <- rep(decay.profile,length(time.ms))
    waveform.df <- data_frame(time.ms,response,decay.setting)
    
    if(is.null(all.traces))
    {
       all.traces <- waveform.df
    }
    else
    {
       all.traces <- bind_rows(all.traces,waveform.df) 
    }
}

```

We now add in some batch and experiment day information

```{r}
all.traces$batch <- ifelse(all.traces$decay.setting > 10,1234,5678)
all.traces$exp.day <- ifelse(all.traces$decay.setting %% 10 == 0,'Day 1','Day 2')
```


```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
print(p)
```



```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response [Panel: batch]'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
p <- p + facet_grid(. ~ batch)
print(p)
```


```{r}
p <- ggplot(data=all.traces,aes(x=time.ms,y=response,colour = as.factor(decay.setting)))
p <- p + geom_point()
p <- p + geom_line()
p <- p + labs(title='Experiment Response [Panel: batch, experiment day]'
                        ,x='Time (ms)', y='Response proportion'
                        ,colour='Decay Setting')
p <- p + facet_grid(exp.day ~ batch)
print(p)
```

Now let's save that out in a .csv format where each row is the trace from an 
experimental row along with the meta data (in this case the decay setting, 
batch and experiment day) from that run.

We have four experimental runs:

+ Day 1
    + Batch 1234
        + Decay setting 20
        
+ Day 1
    + Batch 5678
        + Decay setting 10
        
+ Day 2
    + Batch 1234
        + Decay setting 15
        
+ Day 2
    + Batch 5678
        + Decay setting 5        

So this means we will have four rows in our .csv file.

Let's pull out each run's results into an R object. 

```{r}
d1.b1234.ds20 <- filter(all.traces, exp.day == 'Day 1', batch == 1234, decay.setting == 20)
d1.b5678.ds10 <- filter(all.traces, exp.day == 'Day 1', batch == 5678, decay.setting == 10)

d2.b1234.ds15 <- filter(all.traces, exp.day == 'Day 2', batch == 1234, decay.setting == 15)
d2.b5678.ds05 <- filter(all.traces, exp.day == 'Day 2', batch == 5678, decay.setting == 5)
```

For each of these objects we have 11 rows of 5 variables. We need to flatten this data
into a single row. In actual fact we have 3 bits of meta data (decay setting, 
batch and experiment day) and the 11 reading from the waveform 0 to 100 ms. We are going
to construct 14 column row which we can write out in .csv format.

First we collect the waveform data into a single numeric vector.

```{r}
exp.df <- d1.b1234.ds20
wf <- NULL
num.time.pts <- nrow(exp.df)

for (count in 1:num.time.pts)
{
    wf <- c(wf,exp.df$response[count])
}
```

Now we make another vector with the meta data. We only need to grab the first 
element of the required vector.

```{r}
meta.vec.df <- c(exp.df$exp.day[1],exp.df$batch[1],exp.df$decay.setting[1])
```

Then we can stick these two vectors together and give them names.

```{r}
full.row <- c(meta.vec,wf)
names(full.row) <- c('ExpDay','Batch','DecaySetting'
                        ,'t=0','t=10','t=20','t=30','t=40','t=50'
                        ,'t=60','t=70','t=80','t=90','t=100')

full.row
```
Let's put the above techniques into a function

```{r}
build_WF_Row <- function(exp.df)
{
    wf <- NULL
    num.time.pts <- nrow(exp.df)
    
    for (count in 1:num.time.pts)
    {
        wf <- c(wf,exp.df$response[count])
    }
    
    meta.vec <- c(exp.df$exp.day[1],exp.df$batch[1],exp.df$decay.setting[1])

    full.row <- c(meta.vec,wf)

    full.row.df <- data.frame(lapply(full.row, type.convert), stringsAsFactors=FALSE)
    names(full.row.df) <- c('ExpDay','Batch','DecaySetting'
                        ,'t=0','t=10','t=20','t=30','t=40','t=50'
                        ,'t=60','t=70','t=80','t=90','t=100')
    return(full.row.df)
}
```


Now we'll build up a 4 row object with the data from the experiment. 

```{r}
full.exp.list <- list()
full.exp.list[[1]] <- build_WF_Row(d1.b1234.ds20)
full.exp.list[[2]] <- build_WF_Row(d1.b5678.ds10)
full.exp.list[[3]] <- build_WF_Row(d2.b1234.ds15)
full.exp.list[[4]] <- build_WF_Row(d2.b5678.ds05)
```
We have a list  but what we want it a data_frame. Use rbindlist from th data.table package

```{r}
full.exp.df <- rbindlist(full.exp.list)
full.exp.df$ExpDay <- as.character(full.exp.df$ExpDay)
```


Let's check that looks as we expect


```{r}
full.exp.df[,c('ExpDay','Batch','DecaySetting','t=0','t=10','t=90','t=100')]
```

We can now write this out to a .csv file so that you can prove to yourself
that it is what you would expect.

I'll assume that you current directory is the project dir.

```{r}
exp.file.name <- 'ExperimentFullData.csv'
write.csv(full.exp.df,file = exp.file.name. row.names = FALSE )
```






